---
title: "Airline Project"
author: "ZJ"
date: "10/15/2021"
output: html_document
---

```{r}
library(tidyr)
library(ggplot2)
library(ggmap)
library(leaflet)
library(leaflet.extras)
library(tidyverse)
library(lubridate)
library(scales)
require(gridExtra)
```

```{r}
data <- read.csv("flights1.csv")
airlines <- read.csv("airlines.csv")
airports <- read.csv("airports.csv")
```

### Data Cleaning

```{r}
data$date <- paste(data$YEAR,data$MONTH,data$DAY,sep = "-")

data$SCHEDULED_DEPARTURE1 <- substr(as.POSIXct(sprintf("%04.0f", data$SCHEDULED_DEPARTURE), format='%H%M'), 12, 19)
data$SCHEDULED_DEPARTURE_TIME <- strptime(paste(data$date,data$SCHEDULED_DEPARTURE1),format='%Y-%m-%d %H:%M:%S')

data$DEPARTURE_TIME1 <- substr(as.POSIXct(sprintf("%04.0f", data$DEPARTURE_TIME), format='%H%M'), 12, 19)
data$ACTUAL_DEPARTURE_TIME <- strptime(paste(data$date,data$DEPARTURE_TIME1),format='%Y-%m-%d %H:%M:%S')

data$WHEELS_OFF1 <- substr(as.POSIXct(sprintf("%04.0f", data$WHEELS_OFF), format='%H%M'), 12, 19)
data$WHEELS_OFF_TIME <- strptime(paste(data$date,data$WHEELS_OFF1),format='%Y-%m-%d %H:%M:%S')

data$WHEELS_ON1 <- substr(as.POSIXct(sprintf("%04.0f", data$WHEELS_ON), format='%H%M'), 12, 19)
data$WHEELS_ON_TIME <-  strptime(paste(data$date,data$WHEELS_ON1),format='%Y-%m-%d %H:%M:%S')

data$SCHEDULED_ARRIVAL1 <- substr(as.POSIXct(sprintf("%04.0f", data$SCHEDULED_ARRIVAL), format='%H%M'), 12, 19)
data$SCHEDULED_ARRIVAL_TIME <- strptime(paste(data$date,data$SCHEDULED_ARRIVAL1),format='%Y-%m-%d %H:%M:%S')

data$ARRIVAL_TIME1 <- substr(as.POSIXct(sprintf("%04.0f", data$ARRIVAL_TIME), format='%H%M'), 12, 19)
data$ACTUAL_ARRIVAL_TIME <- strptime(paste(data$date,data$ARRIVAL_TIME1),format='%Y-%m-%d %H:%M:%S')

flights <- data

flights_cleaned <- select(flights,-c("YEAR","MONTH","DAY","SCHEDULED_DEPARTURE","DEPARTURE_TIME","WHEELS_OFF","WHEELS_ON","SCHEDULED_ARRIVAL","ARRIVAL_TIME","SCHEDULED_DEPARTURE1","DEPARTURE_TIME1","WHEELS_OFF1","WHEELS_ON1","SCHEDULED_ARRIVAL1","ARRIVAL_TIME1"))

index <- flights_cleaned$SCHEDULED_DEPARTURE_TIME>flights_cleaned$SCHEDULED_ARRIVAL_TIME & !is.na(flights_cleaned$SCHEDULED_ARRIVAL_TIME)
flights_cleaned$SCHEDULED_ARRIVAL_TIME[index] <- flights_cleaned$SCHEDULED_ARRIVAL_TIME[index]+lubridate::days(1)

index <- flights_cleaned$ACTUAL_DEPARTURE_TIME>flights_cleaned$WHEELS_OFF_TIME & !is.na(flights_cleaned$WHEELS_OFF_TIME)
flights_cleaned$WHEELS_OFF_TIME[index] <- flights_cleaned$WHEELS_OFF_TIME[index]+lubridate::days(1)

index <- flights_cleaned$WHEELS_OFF_TIME>flights_cleaned$WHEELS_ON_TIME & !is.na(flights_cleaned$WHEELS_ON_TIME)
flights_cleaned$WHEELS_ON_TIME[index] <- flights_cleaned$WHEELS_ON_TIME[index]+lubridate::days(1)

index <- flights_cleaned$WHEELS_ON_TIME>flights_cleaned$ACTUAL_ARRIVAL_TIME & !is.na(flights_cleaned$ACTUAL_ARRIVAL_TIME)
flights_cleaned$ACTUAL_ARRIVAL_TIME[index] <- flights_cleaned$ACTUAL_ARRIVAL_TIME[index]+lubridate::days(1)

head(flights_cleaned,5)
```

### Which airline is the best?

```{r}
flights_cleaned$delays <- ifelse(flights_cleaned$AIR_SYSTEM_DELAY > 0,1, ifelse(flights_cleaned$LATE_AIRCRAFT_DELAY > 0,1,ifelse(flights_cleaned$AIRLINE_DELAY > 0,1,0)))
flights_cleaned[is.na(flights_cleaned$delays),]$delays <- 0

avrg_delay <- flights_cleaned %>% group_by(AIRLINE) %>% 
  filter(ARRIVAL_DELAY>0) %>% 
  summarise(avrg_delay=mean(ARRIVAL_DELAY)) %>% 
  arrange(desc(.[,2]))

delays <- flights_cleaned %>% group_by(AIRLINE) %>% summarise(delayed=sum(delays))
total_flights <- flights_cleaned %>%  group_by(AIRLINE) %>% summarise(count=n())
merged <- merge(delays,total_flights,on="AIRLINE")

airline_stats <- merge(merged,avrg_delay,by="AIRLINE")
airline_stats$delay_perc <- (airline_stats$delayed/airline_stats$count)*100

head(airline_stats,5)
```

### 1. Average Delay Time (Negative delay time not included)
  
```{r}
avrg_delay_plot <- airline_stats %>%  ggplot(aes(x=reorder(AIRLINE, avrg_delay),y=avrg_delay,fill=AIRLINE)) + 
  geom_bar(stat="identity",width=0.7) +
  labs(y="Average Delay (Minutes)") +
  theme(axis.title.x=element_blank(),legend.position = "none")
```

### 2. Delay Percentage (Only including air system delay, airline delay and late aircraft delay)
```{r}
delay_perc_plot <- airline_stats %>% ggplot(aes(x=reorder(AIRLINE, delay_perc),y=delay_perc, fill=AIRLINE)) + 
  geom_bar(stat="identity",width=0.7) +
  labs(y="Percentage of Delayed Flights") +
  theme(axis.title.x=element_blank(),legend.position = "none")
```

### 3. Total Flights
```{r}
options(scipen = 10000)

total_flights_plot <- airline_stats %>% ggplot(aes(x = reorder(AIRLINE, -count), y = count, fill = AIRLINE)) + 
  geom_bar(stat="identity",width=0.7) +
  labs(y="Total Flights") +
  scale_y_continuous(labels = comma) +
  theme(axis.title.x=element_blank(),legend.position = "none", axis.text.y = element_text(angle = 90,hjust=0.5))
```


### 1, 2, 3 Combined
```{r, fig.width=7,fig.height=8}
grid.arrange(avrg_delay_plot,delay_perc_plot,total_flights_plot,top="Airline Comparison")
```

### Chart 1, 2 and 3 visualised as a scatter plot
```{r, fig.width=10,fig.height=8}
airline_stats %>% ggplot(aes(x=delay_perc,y=avrg_delay,size=count)) + 
  geom_point(aes(color=AIRLINE)) + 
  geom_text(aes(label=AIRLINE,hjust=-0.7,vjust=0.3),size=3.5) +
  labs(title="Airline Comparison",x="Percentage of Flights Delayed",y="Average Delay (Minutes)",size="Total Flights",color="Airline")
```

From the visualisations, we can see that based on the average delay and percentage of flights delayed, HA, AS and US perform the best. However, they have very few total flights. Comparatively, airlines such as WN, DL, OO and AA have a significantly greater number of flights yet still manage to do relatively well.



### Consider peak months

```{r}
flights_cleaned %>% mutate(month=month(date)) %>% group_by(month) %>% summarise(n=n())
```

The busiest months of the year for airlines tend to be June, July and August.

```{r}
flights_cleaned_busy <- flights_cleaned %>% mutate(month=month(date)) %>% filter(month==6|month==7|month==8)

avrg_delay_b <- flights_cleaned_busy %>% group_by(AIRLINE) %>% 
  filter(ARRIVAL_DELAY>0) %>% 
  summarise(avrg_delay=mean(ARRIVAL_DELAY)) %>% 
  arrange(desc(.[,2]))

delays_b <- flights_cleaned_busy %>% group_by(AIRLINE) %>% summarise(delayed=sum(delays))
total_flights_b <- flights_cleaned_busy %>%  group_by(AIRLINE) %>% summarise(count=n())
merged_b <- merge(delays_b,total_flights_b,on="AIRLINE")

airline_stats_b <- merge(merged_b,avrg_delay_b,by="AIRLINE")
airline_stats_b$delay_perc <- (airline_stats_b$delayed/airline_stats_b$count)*100

head(airline_stats_b,5)
```

### 4.Average Delay Time during peak season (Negative delay time not included)
```{r}
avrg_delay_plot_b <- airline_stats_b %>%  ggplot(aes(x=reorder(AIRLINE, avrg_delay),y=avrg_delay,fill=AIRLINE)) +
  geom_bar(stat="identity",width=0.7) +
  labs(y="Average Delay (Minutes)") +
  theme(axis.title.x=element_blank(),legend.position = "none")
```

### 5. Delay Percentage during peak season (Only including air system delay, airline delay and late aircraft delay)
```{r}
delay_perc_plot_b <- airline_stats_b %>% ggplot(aes(x=reorder(AIRLINE, delay_perc),y=delay_perc, fill=AIRLINE)) +
  geom_bar(stat="identity",width=0.7) +
  labs(y="Percentage of Delayed Flights") +
  theme(axis.title.x=element_blank(),legend.position = "none")
```

### 6. Total Flights during peak season
```{r}
total_flights_plot_b <- airline_stats_b %>% ggplot(aes(x = reorder(AIRLINE, -count), y = count, fill = AIRLINE)) + 
  geom_bar(stat="identity",width=0.7) +
  labs(y="Total Flights") +
  scale_y_continuous(labels = comma) +
  theme(axis.title.x=element_blank(),legend.position = "none",axis.text.y = element_text(angle = 90,hjust=0.5))
```

### 4, 5, 6 Combined
```{r, fig.width=7,fig.height=8}
grid.arrange(avrg_delay_plot_b,delay_perc_plot_b,total_flights_plot_b,top="Airline Comparison (Peak Months)")
```

### Chart 4, 5 and 6 visualised as a scatter plot during peak season
```{r, fig.width=10,fig.height=8}
airline_stats_b %>% ggplot(aes(x=delay_perc,y=avrg_delay,size=count)) + 
  geom_point(aes(color=AIRLINE)) + 
  geom_text(aes(label=AIRLINE,hjust=-0.7,vjust=0.3),size=3.5) +
  labs(title="Airline Comparison (Peak Months)",x="Percentage of Flights Delayed",y="Average Delay (Minutes)",size="Total Flights",color="Airline")
```

When we analyse the performance of the airlines during the busiest months of the year (Jun, Jul, Aug), we find that HA and AS still have the best performance. However, for the larger airlines, DL and OO are still able to main their performance in terms average delay and percentage of delayed flights. AA and WN's performance, on the other hand, visibly deteriorates during the peak months.


### Which airport is best?
```{r}
avrg_airport_delay <- flights_cleaned %>% 
  mutate(airport_delay_duration=ifelse(flights_cleaned$SECURITY_DELAY > 0,flights_cleaned$SECURITY_DELAY, ifelse(flights_cleaned$WEATHER_DELAY > 0,flights_cleaned$WEATHER_DELAY,NA))) %>% 
  group_by(ORIGIN_AIRPORT) %>% 
  filter(airport_delay_duration > 0) %>% 
  summarise(avrg_airport_delay=mean(airport_delay_duration)) %>% 
  arrange(desc(.[,2]))

airports_df <- flights_cleaned %>% group_by(ORIGIN_AIRPORT) %>% summarise(origin=n())

flights_cleaned$airport_delays <- 0
index <- (flights_cleaned$SECURITY_DELAY > 0 | flights_cleaned$WEATHER_DELAY > 0) & !is.na(flights_cleaned$SECURITY_DELAY)
flights_cleaned$airport_delays[index] <- flights_cleaned$airport_delays[index] + 1
airport_delays <- flights_cleaned %>% group_by(ORIGIN_AIRPORT) %>% summarise(airport_delays=sum(airport_delays))

taxi_out_time <- flights_cleaned %>% .[!is.na(flights_cleaned$TAXI_OUT),] %>% group_by(ORIGIN_AIRPORT) %>% summarise(avrg_taxi_out=mean(TAXI_OUT))
taxi_in_time <- flights_cleaned %>% .[!is.na(flights_cleaned$TAXI_IN),] %>% group_by(DESTINATION_AIRPORT) %>% summarise(avrg_taxi_in=mean(TAXI_IN))

merged2 <- merge(x=airports_df,y=airport_delays,by="ORIGIN_AIRPORT",all.x=T)
merged2 <- merge(merged2,taxi_out_time,by="ORIGIN_AIRPORT",all.x=T)
merged2 <- merge(x=merged2,y=taxi_in_time,by.x="ORIGIN_AIRPORT",by.y="DESTINATION_AIRPORT",all.x=T)
merged2 <- merge(x=merged2,y=avrg_airport_delay,by="ORIGIN_AIRPORT",all.x=T)
merged2[is.na(merged2$avrg_airport_delay),]$avrg_airport_delay <- 0
merged2$delay_perc <- (merged2$airport_delays/merged2$origin)*100

head(merged2,5)
```

### Delay % against Average Airport Delay
```{r, fig.width=10.5,fig.height=8}
merged2 %>% filter(origin>=1000) %>%     
  ggplot(aes(x=delay_perc,y=avrg_airport_delay,color=origin,size=origin)) + geom_point(alpha=0.7) +
  labs(title="Airport Comparison",x="Percentage of Flights Delayed",y="Average Delay (Minutes)",color="Total Flights",size="Total Flights") +
  scale_color_gradient(low="blue", high="red") +
  geom_text(aes(label=ORIGIN_AIRPORT,hjust=-0.5,vjust=0.5),size=3)
```

For this visualisation, we filter out airports with fewer than 1000 total flights. Taking into account the total flights at each airport, LAX and SFO perform the best with the lowest average delay and lowest percentage of flights delayed.

### Average Taxi Out against Average Taxi In
```{r, fig.width=10,fig.height=8}
merged2 %>% filter(origin>=1000) %>%                                               
  ggplot(aes(x=avrg_taxi_out,y=avrg_taxi_in,color=origin,size=origin)) + geom_point(alpha=0.7) + 
  labs(title="Airport Comparison",x="Average Taxi Out (Minutes)",y="Average Taxi In (Minutes)",color="Total Flights",size="Total Flights") +
  scale_color_gradient(low="blue", high="red") +
  geom_text(aes(label=ORIGIN_AIRPORT,hjust=0.5,vjust=-1.4),size=3)
```

Similarly, here, we also filter out airports with fewer than 1000 total flights. When it comes to average taxi in and out duration, LAX performs quite poorly compared to others, despite doing well in the terms of delays. Airports such as PHX, SFO and LAS have better average taxi in and out durations.

### Airport connectivity
```{r}
flights_cleaned %>% group_by(ORIGIN_AIRPORT) %>% 
  summarise(count = n_distinct(DESTINATION_AIRPORT)) %>% 
  arrange(-count) %>% 
  top_n(20) %>% 
  ggplot(aes(x=reorder(ORIGIN_AIRPORT, count),y=count,fill=ORIGIN_AIRPORT)) + 
  geom_bar(stat="identity",width=0.7) + 
  labs(title="Airport Connectivity",y="Number of Unique Routes") +
  theme(axis.title.y=element_blank(),legend.position = "none") +
  coord_flip()
```

Airports are also measured by their connectivity, which is the number of connections an airport has to other airports. In this regard, ATL, ORD, DFW, 10397 and 13930 are the top 5.


### What is the seasonal effect on flight performance?
```{r}
num_flights <- flights_cleaned %>% mutate(month=month(date)) %>% group_by(month) %>% summarise(num_flights=n())

delay_types <- flights_cleaned %>% select(AIR_SYSTEM_DELAY,SECURITY_DELAY,AIRLINE_DELAY,LATE_AIRCRAFT_DELAY,WEATHER_DELAY,date)
delay_types <- delay_types[complete.cases(delay_types),]
delay_types <- delay_types %>% mutate(month=month(date)) %>% 
  group_by(month) %>% 
  summarise(AIR_SYSTEM_DELAYS=sum(AIR_SYSTEM_DELAY!=0),
            SECURITY_DELAYS=sum(SECURITY_DELAY!=0),
            AIRLINE_DELAYS=sum(AIRLINE_DELAY!=0),
            LATE_AIRCRAFT_DELAYS=sum(LATE_AIRCRAFT_DELAY!=0),
            WEATHER_DELAYS=sum(WEATHER_DELAY!=0)) %>% 
  merge(num_flights,by="month") %>% 
  mutate(AIR_SYSTEM_DELAYS_PERC=(AIR_SYSTEM_DELAYS/num_flights)*100,
         SECURITY_DELAYS_PERC=(SECURITY_DELAYS/num_flights)*100,
         AIRLINE_DELAYS_PERC=(AIRLINE_DELAYS/num_flights)*100,
         LATE_AIRCRAFT_DELAYS_PERC=(LATE_AIRCRAFT_DELAYS/num_flights)*100,
         WEATHER_DELAYS_PERC=(WEATHER_DELAYS/num_flights)*100)

head(delay_types,5)
```

### Delay types across months
```{r}
delay_types %>% ggplot() + 
  geom_line(aes(x=month,y=AIR_SYSTEM_DELAYS_PERC,color="Air System Delays")) + geom_point(aes(x=month,y=AIR_SYSTEM_DELAYS_PERC,color="Air System Delays")) + 
  geom_line(aes(x=month,y=SECURITY_DELAYS_PERC,color="Security Delays")) + geom_point(aes(x=month,y=SECURITY_DELAYS_PERC,color="Security Delays")) +
  geom_line(aes(x=month,y=AIRLINE_DELAYS_PERC,color="Airline Delays")) + geom_point(aes(x=month,y=AIRLINE_DELAYS_PERC,color="Airline Delays")) +
  geom_line(aes(x=month,y=LATE_AIRCRAFT_DELAYS_PERC,color="Late Aircraft Delays")) + geom_point(aes(x=month,y=LATE_AIRCRAFT_DELAYS_PERC,color="Late Aircraft Delays")) +
  geom_line(aes(x=month,y=WEATHER_DELAYS_PERC,color="Weather Delays")) + geom_point(aes(x=month,y=WEATHER_DELAYS_PERC,color="Weather Delays")) +
  labs(title="Seasonal Effect on Various Types of Delays",y="Delay Percentage",colour="Type of Delay") +
  theme(axis.title.x=element_blank()) +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12),labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```

From the visualisation, we can see that the periods with the greatest effects on delays are Jun-Jul and Dec-Feb. These periods experience the highest percentage of delays.

### Taxi In and Out across the year
```{r}
taxi_plot <- flights_cleaned %>% .[complete.cases(flights_cleaned$TAXI_OUT),] %>% 
  .[complete.cases(.$TAXI_IN),] %>% 
  mutate(month=month(date)) %>% 
  group_by(month) %>% 
  summarise(avrg_taxi_in=mean(TAXI_IN),avrg_taxi_out=mean(TAXI_OUT))
```

### Taxi In Plot  
```{r}
taxi_in <- taxi_plot %>% ggplot() + geom_line(aes(x=month,y=avrg_taxi_in)) + 
  geom_point(aes(x=month,y=avrg_taxi_in)) +
  labs(y="Average Taxi In (Minutes)") +
  theme(axis.title.x=element_blank()) +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12),labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```

### Taxi Out Plot
```{r}
taxi_out <- taxi_plot %>% ggplot() + geom_line(aes(x=month,y=avrg_taxi_out)) + 
  geom_point(aes(x=month,y=avrg_taxi_out)) +
  labs(y="Average Taxi Out (Minutes)") +
  theme(axis.title.x=element_blank()) +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12),labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```

### Cancelled Flights
```{r}
cancelled_flights <- flights_cleaned %>% mutate(month=month(date)) %>% 
  group_by(month) %>% 
  summarise(cancelled_flights=sum(CANCELLED)) %>% 
  ggplot() + geom_line(aes(x=month,y=cancelled_flights)) + 
  geom_point(aes(x=month,y=cancelled_flights)) +
  labs(y="Cancelled Flights") +
  theme(axis.title.x=element_blank()) +
  scale_x_continuous(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12),labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))
```

### Combine Taxi In, Taxi Out and Cancelled Flights
```{r, fig.width=8,fig.height=7}
grid.arrange(taxi_in,taxi_out,cancelled_flights,top="Seasonal Effect on Performance")
```

When comparing taxi durations, we find a similar trend, with Dec-Feb and Jun-Aug experiencing the highest average taxi in and out durations. The number of cancelled flights is also high in Jan, Feb, Jun and Dec. Notably, Feb has the highest number of cancellations by far.


### Visualise data on a map
```{r}
ggmap::register_google(key = 'AIzaSyDRyVUxMAnQuuKaqzXI2j4F9US535l4TmI')

connectivity <- flights_cleaned %>% group_by(ORIGIN_AIRPORT) %>% 
  summarise(unique_routes = n_distinct(DESTINATION_AIRPORT))

airports_cleaned <- airports[complete.cases(airports),]
airport_connectivity <- merge(x=airports_cleaned,y=connectivity,by.x="IATA_CODE",by.y="ORIGIN_AIRPORT")

airport_connectivity %>% arrange(desc(unique_routes)) %>%
  select(IATA_CODE,AIRPORT,LATITUDE,LONGITUDE,unique_routes) %>% 
  top_n(10)
```

```{r}
loc <-c(lon=-123,lat=50)

map1 <- get_map(location=loc,zoom=2,source="google",maptype="roadmap")
m1 <- ggmap(map1, base_layer = ggplot(data=airport_connectivity,aes(x=LONGITUDE,y=LATITUDE,color=unique_routes)))
```

```{r, fig.width=8,fig.height=8}
# Map of Airport Connectivity
m1 + geom_point(alpha=0.9,size=1.5) +
  geom_point(shape = 1,size = 1.5,colour = "black",alpha=0.2) +
  labs(title="Airport Connectivity",x="Longitude",y="Latitude",color="Unique Routes") +
  scale_x_continuous(limits = c(-180, -60), expand = c(0, 0)) +
  scale_y_continuous(limits = c(10, 75), expand = c(0, 0)) +
  scale_color_gradient(low="blue", high="red")
```

This visualisation shows the airports with the best connectivity.

### Regions that experience the most weather delays
```{r}
weather_delay <- flights_cleaned %>% filter(WEATHER_DELAY>0) %>% 
  select(ORIGIN_AIRPORT,WEATHER_DELAY) %>% 
  merge(x=.,y=airports,by.x="ORIGIN_AIRPORT",by.y="IATA_CODE") %>% 
  .[complete.cases(.),]

leaflet(weather_delay) %>% addProviderTiles(providers$CartoDB.DarkMatter) %>% 
  addWebGLHeatmap(lng=~LONGITUDE, lat=~LATITUDE, size = 200000)
```

This visualisation shows the regions which experience the most weather delays.
